---
title: Estimating the metabolic index and its threshold effects within a species distribution model with observation data
output: 
  github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/README-",
  out.width = "50%",
  cache = FALSE,
  dpi = 72
)
```

This vignette demonstrates using an optional estimation feature within the sdmTMB R package developed and used in Indivero et al. (in prep). This feature is specifically designed for using spatial data on oxygen, temperature, and species density to estimate:

1. The metabolic index, with an estimated parameter Eo.
2. A threshold function of the effect of the metabolic index on fish density (with estimated parameters $s50$, $s95$, and scaling parameter $\psi$ (called `smax` in sdmTMB)).

This is done within the estimation model of a generalized linear mixed model, i.e. simultaneously estimating covariates effects and spatial random effects. 

For further details on the equations of the metabolic index and the form of the threshold function, see Indivero et al. (in prep).

Here, we will use as an example the data for sablefish (*Anoplopoma fimbria*) from Indivero et al. This data is from 2010--2015 of the National Oceanic and Atmospheric Administration West Coast Bottom Trawl Survey, [Keller et al. 2017](https://repository.library.noaa.gov/view/noaa/14179), spanning Washington, Oregon, and California of the California Current. 

## Downloading the branch

Currently, this feature is included in a separate branch of sdmTMB, called `sdmTMB-newlogistic`. We plan to merge it into the main branch soon. To install the package in R:

```{r install package, echo=TRUE, eval=FALSE}
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE, ref = "newlogistic")
```

```{r, eval=TRUE}
library(sdmTMB)
library(ggplot2)
library(dplyr)
```

## Data Inputs

Load the example data:

```{r data, echo=TRUE}
dat <- readRDS("example_data.rds")
```

This model feature requires three columns in the data frame provided to `sdmTMB()` (in addition to latitude and longitude, if a spatial model):

1. **species density** (e.g., catch-per-unit-effort, or other response variable) (column can be named any valid column name)
2. **oxygen** (must be in kiloPascals (kPa), and column must be named `po2`
3. **temperature** (must be in inverse temperature and Kelvin, see below) and column must be named `invtemp`. 

Any additional time variable or environmental covariates can be included in the data at the same coordinates as the catch, temperature, and oxygen data.

### Oxygen:

Oxygen needs to be in saturation of kPa. For example and functions of how to convert from mL per L to kPa, see [util_funs.R](https://github.com/jindivero/estimating_mi_from_distribution2/blob/f696e18bb8bf5b20d2127924eff033822d6fc2e4/Code/util_funs.R).

### Inverse temperature

Temperature needs to be inverse temperature, in Kelvin, and in reference to a set temperature and corrected with Boltzmann's constant in an Arrhenius equation. Below is an example of converting temperature in Celsius to the correct inverse temperature format.

```{r inverse temperature, echo=TRUE}
# Constants
kelvin <- 273.15
boltz <- 0.000086173324
tref <- 12
# Calculate inverse temp
dat$invtemp <- (1 / boltz) * (1 / (dat$temp + 273.15) - 1 / (tref + 273.15))
```

### Species density

In this example, species density is catch-per-unit-effort, in biomass kg km^-2^. We have not tested with other response variables, though in theory this model could work with other measurements of density (e.g., counts) or other response variables (e.g., size, condition, etc.).

### Other covariates

The metabolic index threshold feature only requires oxygen, temperature, and species density. However, spatial data, time data (e.g., year), and other environmental covariates can also be included. In this example, we also include year and depth as covariates.

```{r view data, echo=TRUE, include=TRUE, eval=TRUE}
head(dat)
```

## Model Estimation

The Metabolic Index estimation is implemented using the standard `sdmTMB()` (or `sdmTMB_cv()` for cross validation) function calls in sdmTMB while adding ` + logistic(mi)` to the model equation. 
This can be added in combination with most other model structure options in sdmTMB.
It can be used with or without temporal, spatial, and spatio-temporal variation, and with other fixed effects in the model equation. However, it cannot currently be used with another threshold variable or with penalized smoothers.
<!-- TODO: or a smoother SA: is this true!? -->

### Make a mesh for the SPDE

Next we construct a mesh to implement the SPDE approach:

```{r mesh, echo=TRUE}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), n_knots = 250)
```

### Fit model

In this example, we will fit a model of fish density to the metabolic index in addition to year, and a quadratic effect of depth. We will include spatial random variation, but not spatio-temporal variation, and we will use a Tweedie observation distribution with a log link.

<!-- Additionally, to help with estimation, we are going to use starting values and use 2 "Newton loops" (additional optimization calls with a Newton optimizer).  -->

We will not use lower and upper bounds here, but the code below shows where these can be added (see the `Inf` and `-Inf`). Starting values for the threshold parameters can be specified in a matrix that is fed to the `sdmTMB()` function (e.g., as in the `start` matrix in the example below).

Note that there are four estimated parameters of the metabolic index threshold feature (in order: `s50`, `s95`, `smax`, `Eo`), so if you want to use any values for starting or bounds, it needs to be a vector (matrix) of length (rows) four, with `NA`s for any parameter you do not want to include a starting value for.

```{r fit model, echo=TRUE, results='hide'}
# Make starting values
start <- matrix(0, ncol = 1, nrow = 4)
start[1, 1] <- 2 # s50
start[2, 1] <- 2 # s95
start[3, 1] <- 30 # smax
start[4, 1] <- 0.3 # Eo

# Fit model
fit <- sdmTMB(
  cpue_kg_km2 ~ -1 + as.factor(year) + logistic(mi) + log_depth_scaled + log_depth_scaled2,
  data = dat,
  spatial = "on",
  spatiotemporal = "off",
  mesh = mesh,
  family = tweedie(link = "log"),
  silent = FALSE,
  control = sdmTMBcontrol(
    start = list(b_threshold = start),
    lower = list(b_threshold = c(-Inf, -Inf, -Inf, -Inf)),
    upper = list(b_threshold = c(Inf, Inf, Inf, Inf))
  )
)
```

### Adding a prior

A prior can be added using `priors = sdmTMBpriors(threshold = ...)`.
Here, we will add a prior on `Eo` only by leaving the others as `NA`.

```{r prior, echo=TRUE, results='hide'}
fit <- sdmTMB(
  cpue_kg_km2 ~ -1 + as.factor(year) + logistic(mi) + log_depth_scaled + log_depth_scaled2,
  data = dat,
  spatial = "on",
  spatiotemporal = "off",
  mesh = mesh,
  silent = FALSE,
  family = tweedie(link = "log"),
  control = sdmTMBcontrol(
    start = list(b_threshold = start),
    lower = list(b_threshold = c(-Inf, -Inf, -Inf, -Inf)),
    upper = list(b_threshold = c(Inf, Inf, Inf, Inf))
  ),
  priors = sdmTMBpriors(threshold = normal(c(NA, NA, NA, 0.3477), c(NA, NA, NA, 0.1455)))
)
```

## Interpreting Model Output

Standard sdmTMB functions can be used to evaluate the model fit.

First, we can run a 'sanity' check on our model:

```{r}
sanity(fit, gradient_thresh = 0.01)
```

The parameters estimated from the metabolic index and threshold function start with the prefix `mi-` in the parameter list.

NOTE: The `s95` parameter value reported is actually a delta value (difference between `s95` and `s50`).
To calculate the value of `s95` itself (i.e., the value of the metabolic index at which the logged response on fish is a 95% reduction on density), add the reported values: `s50 + s95`.

```{r interpret, echo=TRUE, eval=TRUE, include=TRUE}
# View parameter estimates
summary(fit)

# Get AIC of model fit
AIC(fit)
```

### Marginal effects

We can pull out the parameter estimates for the metabolic index threshold feature to calculate the metabolic index from observed temperature and oxygen and the marginal effect of the metabolic index on fish density.

To evaluate how the response is impacted across the range of estimated Eo values (e.g. the 95% confidence interval of the maximum likelihood estimate) the function can be recreated:

```{r marginal effect, echo=TRUE, eval=TRUE, include=TRUE}
## Extract parameters using tidy
par_estimates <- tidy(fit, conf.int = TRUE, effects = "fixed")
par_estimates_rand <- tidy(fit, conf.int = TRUE, effects = "ran_pars")
par_estimates <- bind_rows(par_estimates, par_estimates_rand)

# Or directly from model object
p <- as.list(fit$sd_report, "Estimate", report = TRUE)
s50 <- as.numeric(p$s50)
delta <- as.numeric(p$s95)
smax <- as.numeric(p$s_max)
Eo <- as.numeric(p$Eo)

## Calculate metabolic index
dat$metabolic_index <- dat$po2 * exp(Eo * dat$invtemp)

## Calculate marginal effect of threshold function
# Re-arrange parameters
a <- log(smax / (log(0.5) + smax) - 1)
b <- log(smax / (log(0.95) + smax) - 1)
beta0 <- -a + s50 * (b - a) / delta
beta1 <- (a - b) / delta
# Calculate
dat$effect <- exp(smax * (1 / (1 + exp(-beta0 - beta1 * dat$metabolic_index)) - 1))

## Plot
ggplot(dat, aes(x = metabolic_index, y = effect)) +
  geom_point()
```

### Threshold oxygen level across temperatures

The parameter estimates can also be used to calculate the values of oxygen across a range of temperatures that correspond to the limiting metabolic index threshold (either s50 or s95).

```{r threshold, echo=TRUE, eval=TRUE, include=TRUE}
# For the s50 threshold
dat$po2_s50 <- s50 / exp(Eo * dat$invtemp)
# For the s95 threshold
s95 <- s50 + delta
dat$po2_s95 <- s95 / exp(Eo * dat$invtemp)

ggplot(dat, aes(x = invtemp, y = po2_s50)) +
  geom_point()

# Or with temperature in Celsius
ggplot(dat, aes(x = temp, y = po2_s50)) +
  geom_point()
```
